---
title: 'Johns Hopkins Especialization'
output:
  html_document:
  keep_md: yes
---
#Reproducible Research: Peer Assessment 1

## Loading and preprocessing the data
```{r, echo = TRUE}
library(dplyr)
library(ggplot2)
library(ggpubr)

MyData <- read.csv("activity.csv", header=TRUE, stringsAsFactors = FALSE, sep=",")
print ("Data loaded with read.csv")
```
## What is mean total number of steps taken per day?
```{r, echo = TRUE}
NewGroup <- group_by(MyData, date)
NewGroup <- summarise(NewGroup, stepsPerDay = sum(steps, na.rm=TRUE), intervalsPerDay = n())
print("Steps per day")

ggplot(NewGroup, aes(x = date, y = stepsPerDay, color = I("#FF4081"))) + geom_bar(stat = "identity", fill = I("#512DA8"))

meanA <- mean(NewGroup$stepsPerDay, na.rm = TRUE)
medianA <- median(NewGroup$stepsPerDay, na.rm = TRUE)

print(paste("Mean of steps per day", meanA))
print(paste("Median of steps per day", medianA))

pA <- ggplot(NewGroup, aes(stepsPerDay, color = I("#FF4081")) ) + geom_histogram(fill = I("#512DA8"))
print (pA)

```

## What is the average daily activity pattern?

```{r, echo = TRUE}

NewGroup <- group_by(MyData, interval)
NewGroup <- summarise(NewGroup, stepsPerInterval = sum(steps, na.rm=TRUE), daysPerInterval = n())

maxValue <- max(NewGroup$stepsPerInterval, na.rm = TRUE)
maxInterval <- subset(NewGroup, stepsPerInterval == maxValue)

ggplot(NewGroup, 
       aes(x = interval, y = stepsPerInterval, group = 1, color = I("red"))) + 
       geom_line() + 
       annotate("text", x = maxInterval$interval - 210, y = maxInterval$stepsPerInterval, label = "Max Value >") +
       annotate("point", shape = 21, size = 3, x = maxInterval$interval, y = maxInterval$stepsPerInterval, fill = I("#795548"))

print(paste("The interval with more steps is the number", maxInterval$interval, "with", maxValue, "steps taken"))

```

## Imputing missing values

```{r, echo = TRUE}
totalNAs <- sum(is.na(MyData$steps))
totalRows <- nrow(MyData)
percetagofNAs <- totalNAs / totalRows

print(paste("Total of NAs in steps column:", totalNAs))
print(paste("Percentage of NAs in step column:", percetagofNAs))

```

### Defining strategy to replace NAs.  
I will apply four strategies:

1. using the value 0
2. using the mean of that 5-minute interval
3. using the madian ot that 5-minute interval

### Using the value 0

```{r, echo = TRUE}

newData <- mutate(MyData, steps = ifelse(is.na(steps), 0, steps))

NewGroup <- group_by(newData, date)
NewGroup <- summarise(NewGroup, stepsPerDay = sum(steps, na.rm=TRUE), intervalsPerDay = n())
print("Steps per day")

mean1 <- mean(NewGroup$stepsPerDay, na.rm = TRUE)
median1 <- median(NewGroup$stepsPerDay, na.rm = TRUE)

header <- c("Measure", "Original Dataset", "Using 0 instead of NAs")
row1 <- c("mean", meanA, mean1)              
row2 <- c("median", medianA, mean1)

a <- data.frame(header, row1, row2)
print(a)                  

print("Do these values differ from the estimates from the first part of the assignment?")
print("As we can see the value are the same.")

p1 <- ggplot(NewGroup, aes(stepsPerDay, color = I("#FF4081")) ) + geom_histogram(fill = I("#512DA8"))

ggarrange(pA, p1,
          labels = c("With NAs", "With 0 in that 5-minute interval"), 
          ncol = 2, nrow = 1)

print("What is the impact of imputing missing data on the estimates of the total daily number of steps?")
print("In this case nothing happens.")

```

### Using the mean of that 5-minute interval

```{r, echo = TRUE}

NewGroup <- group_by(MyData, interval)
NewGroup <- summarise(NewGroup, stepsPerInterval = sum(steps, na.rm=TRUE), daysPerInterval = n(), medianPerInterval = median(steps, na.rm=TRUE), meanPerInterval = mean(steps, na.rm=TRUE))

newData <- merge(MyData, NewGroup, by.x = "interval", by.y = "interval")
newData <- mutate(newData, steps = ifelse(is.na(steps), meanPerInterval, steps))

NewGroup <- group_by(newData, date)
NewGroup <- summarise(NewGroup, stepsPerDay = sum(steps, na.rm=TRUE), intervalsPerDay = n())
print("Steps per day")

mean1 <- mean(NewGroup$stepsPerDay, na.rm = TRUE)
median1 <- median(NewGroup$stepsPerDay, na.rm = TRUE)

header <- c("Measure", "Original Dataset", "Using mean instead of NAs")
row1 <- c("mean", meanA, mean1)              
row2 <- c("median", medianA, mean1)

a <- data.frame(header, row1, row2)
print(a)                  

print("Do these values differ from the estimates from the first part of the assignment?")
print("As we can see the value are different.")

p1 <- ggplot(NewGroup, aes(stepsPerDay, color = I("#FF4081")) ) + geom_histogram(fill = I("#512DA8"))

ggarrange(pA, p1, 
          labels = c("With NAs", "With mean of that 5-minute interval"),
          ncol = 2, nrow = 1)

print("What is the impact of imputing missing data on the estimates of the total daily number of steps?")
print("In this case we can aprreciate that the graphic changes.")

```

### Using the median of that 5-minute interval

```{r, echo = TRUE}

NewGroup <- group_by(MyData, interval)
NewGroup <- summarise(NewGroup, stepsPerInterval = sum(steps, na.rm=TRUE), daysPerInterval = n(), medianPerInterval = median(steps, na.rm=TRUE), meanPerInterval = mean(steps, na.rm=TRUE))

newData <- merge(MyData, NewGroup, by.x = "interval", by.y = "interval")
newData <- mutate(newData, steps = ifelse(is.na(steps), medianPerInterval, steps))

NewGroup <- group_by(newData, date)
NewGroup <- summarise(NewGroup, stepsPerDay = sum(steps, na.rm=TRUE), intervalsPerDay = n())
print("Steps per day")

mean1 <- mean(NewGroup$stepsPerDay, na.rm = TRUE)
median1 <- median(NewGroup$stepsPerDay, na.rm = TRUE)

header <- c("Measure", "Original Dataset", "Using median instead of NAs")
row1 <- c("mean", meanA, mean1)              
row2 <- c("median", medianA, mean1)

a <- data.frame(header, row1, row2)
print(a)                  

print("Do these values differ from the estimates from the first part of the assignment?")
print("As we can see the values also are different.")

p1 <- ggplot(NewGroup, aes(stepsPerDay, color = I("#FF4081")) ) + geom_histogram(fill = I("#512DA8"))

ggarrange(pA, p1, 
          labels = c("With NAs", "With median of that 5-minute interval"),
          ncol = 2, nrow = 1)

print("What is the impact of imputing missing data on the estimates of the total daily number of steps?")
print("In this case we can aprreciate that the graphic also changes.")

```

## Are there difference

### Using the value 0
### Creating factor for weekday and weekend

```{r, echo = TRUE}

newData <- mutate(MyData, steps = ifelse(is.na(steps), 0, steps))
newData$typeDay <- ifelse(weekdays(as.Date(newData$date)) %in% c("sábado", "domingo"), "weekend", "weekday")

print(head(newData, 10))
print(table(newData$typeDay))

```

### Comparing weekday and weekend

```{r, echo = TRUE}

NewGroup <- select(newData, interval, steps, typeDay)
NewGroup <- filter(NewGroup, typeDay == "weekend")
NewGroup <- group_by(NewGroup, interval)
NewGroup <- summarise(NewGroup, stepsPerInterval = sum(steps, na.rm=TRUE), daysPerInterval = n())

p1 <- ggplot(NewGroup, 
       aes(x = interval, y = stepsPerInterval, group = 1, color = I("red"))) + 
       geom_line() + ylim(0, 12000) + ggtitle("weekend")

NewGroup <- select(newData, interval, steps, typeDay)
NewGroup <- filter(NewGroup, typeDay == "weekday")
NewGroup <- group_by(NewGroup, interval)
NewGroup <- summarise(NewGroup, stepsPerInterval = sum(steps, na.rm=TRUE), daysPerInterval = n())

p2 <- ggplot(NewGroup, 
       aes(x = interval, y = stepsPerInterval, group = 1, color = I("red"))) + 
       geom_line() + ylim(0, 12000) + ggtitle("weekday")

ggarrange(p1, p2, ncol = 1, nrow = 2)
w1 <- p1
w2 <- p2
```

### Using the mean of that 5-minute interval
### Creating factor for weekday and weekend

```{r, echo = TRUE}

NewGroup <- group_by(MyData, interval)
NewGroup <- summarise(NewGroup, stepsPerInterval = sum(steps, na.rm=TRUE), daysPerInterval = n(), medianPerInterval = median(steps, na.rm=TRUE), meanPerInterval = mean(steps, na.rm=TRUE))

newData <- merge(MyData, NewGroup, by.x = "interval", by.y = "interval")
newData <- mutate(newData, steps = ifelse(is.na(steps), meanPerInterval, steps))
newData$typeDay <- ifelse(weekdays(as.Date(newData$date)) %in% c("sábado", "domingo"), "weekend", "weekday")

print(head(newData, 10))
print(table(newData$typeDay))

```

### Comparing weekday and weekend

```{r, echo = TRUE}

NewGroup <- select(newData, interval, steps, typeDay)
NewGroup <- filter(NewGroup, typeDay == "weekend")
NewGroup <- group_by(NewGroup, interval)
NewGroup <- summarise(NewGroup, stepsPerInterval = sum(steps, na.rm=TRUE), daysPerInterval = n())

p1 <- ggplot(NewGroup, 
       aes(x = interval, y = stepsPerInterval, group = 1, color = I("red"))) + 
       geom_line() + ylim(0, 12000) + ggtitle("weekend")

NewGroup <- select(newData, interval, steps, typeDay)
NewGroup <- filter(NewGroup, typeDay == "weekday")
NewGroup <- group_by(NewGroup, interval)
NewGroup <- summarise(NewGroup, stepsPerInterval = sum(steps, na.rm=TRUE), daysPerInterval = n())

p2 <- ggplot(NewGroup, 
       aes(x = interval, y = stepsPerInterval, group = 1, color = I("red"))) + 
       geom_line() + ylim(0, 12000) + ggtitle("weekday")

ggarrange(p1, p2, ncol = 1, nrow = 2)
w3 <- p1
w4 <- p2

```

### Using the median of that 5-minute interval
### Creating factor for weekday and weekend

```{r, echo = TRUE}

NewGroup <- group_by(MyData, interval)
NewGroup <- summarise(NewGroup, stepsPerInterval = sum(steps, na.rm=TRUE), daysPerInterval = n(), medianPerInterval = median(steps, na.rm=TRUE), meanPerInterval = mean(steps, na.rm=TRUE))

newData <- merge(MyData, NewGroup, by.x = "interval", by.y = "interval")
newData <- mutate(newData, steps = ifelse(is.na(steps), medianPerInterval, steps))
newData$typeDay <- ifelse(weekdays(as.Date(newData$date)) %in% c("sábado", "domingo"), "weekend", "weekday")

print(head(newData, 10))
print(table(newData$typeDay))

```

### Comparing weekday and weekend

```{r, echo = TRUE}

NewGroup <- select(newData, interval, steps, typeDay)
NewGroup <- filter(NewGroup, typeDay == "weekend")
NewGroup <- group_by(NewGroup, interval)
NewGroup <- summarise(NewGroup, stepsPerInterval = sum(steps, na.rm=TRUE), daysPerInterval = n())

p1 <- ggplot(NewGroup, 
       aes(x = interval, y = stepsPerInterval, group = 1, color = I("red"))) + 
       geom_line() + ylim(0, 12000) + ggtitle("weekend")

NewGroup <- select(newData, interval, steps, typeDay)
NewGroup <- filter(NewGroup, typeDay == "weekday")
NewGroup <- group_by(NewGroup, interval)
NewGroup <- summarise(NewGroup, stepsPerInterval = sum(steps, na.rm=TRUE), daysPerInterval = n())

p2 <- ggplot(NewGroup, 
       aes(x = interval, y = stepsPerInterval, group = 1, color = I("red"))) + 
       geom_line() + ylim(0, 12000) + ggtitle("weekday")

ggarrange(p1, p2, ncol = 1, nrow = 2)
w5 <- p1
w6 <- p2

ggarrange(w1, w2, w3, w4, w5, w6, ncol = 2, nrow = 3)

```

End of this report